---
title: "Computations of Stocastic Processes"
Author: Aubrey Undi Phiri 
output: html_document
date: "2025-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Note that this document was only used for compputations, the explanations of all the processes are in the pdf file


```{r}


# Load necessary package for matrix exponentiation
library(expm) #allow raising matrix to a power
library(ggplot2)
library(gganimate)
library(dplyr)
library(av)  # Load av for video rendering

```



Question 1
b
```{r}


# Define the transition matrix P
P <- matrix(c(
    1,   0,   0,   0,   0,   0,   0,   0,   0,
    0.4, 0,   0.6, 0,   0,   0,   0,   0,   0,
    0,   0.4, 0,   0.6, 0,   0,   0,   0,   0,
    0,   0,   0.4, 0,   0.6, 0,   0,   0,   0,
    0,   0,   0,   0.4, 0,   0.6, 0,   0,   0,
    0,   0,   0,   0,   0.4, 0,   0.6, 0,   0,
    0,   0,   0,   0,   0,   0.4, 0,   0.6, 0,
    0,   0,   0,   0,   0,   0,   0.4, 0,   0.6,
    0,   0,   0,   0,   0,   0,   0,   0,   1
), nrow = 9, byrow = TRUE)

# Compute P^4 using matrix exponentiation
P_4 <- P %^% 4

# Label the rows and columns from 0 to 8
colnames(P_4) <- 0:8
rownames(P_4) <- 0:8

# Print the result
print(P_4)

```

Question 1
c)

The code below outputs a short video showing the simulation of the Markov chain.

```{r}
#libraries that are needed:
# Load libraries
#library(ggplot2)
#library(gganimate)
#library(dplyr)
#library(av)  # Load av for video rendering

# Set seed and define transition matrix
set.seed(123)
transition_matrix <- matrix(c(0.1279, 0.8721, 0.6631, 0.3369), nrow = 2, byrow = TRUE)
states <- c("V", "C")

# Simulate Markov chain
n_steps <- 20
sequence <- tibble(Step = 1:n_steps, State = character(n_steps))
sequence$State[1] <- "V"
for (i in 2:n_steps) {
  sequence$State[i] <- sample(states, size = 1, prob = transition_matrix[sequence$State[i - 1] == states, ])
}
sequence <- sequence %>% mutate(StateNum = ifelse(State == "V", 1, 2))

# Plot and animate
p <- ggplot(sequence, aes(x = Step, y = StateNum, color = State, group = 1)) +
  geom_point(size = 5) +
  geom_line(color = "black", linewidth = 1) +  # Explicitly set the group aesthetic
  scale_color_manual(values = c("V" = "blue", "C" = "red")) +
  scale_y_continuous(breaks = 1:2, labels = states) +
  theme_minimal() +
  labs(title = "Markov Chain Simulation of State Transitions",
       subtitle = "Step: {frame_time}",
       x = "Time Step",
       y = "State") +
  transition_reveal(Step)

# Save and show animation as video (MP4 format)
animate(p, nframes = n_steps, fps = 2, renderer = av_renderer("markov_simulation.mp4"))


```

Question 1

c) The full image of the simulation

```{r}

# Simulation settings
set.seed(123)
nTrials <- 1  # Number of trials
nSteps <- 20    # Number of steps
states <- c("V", "C")  # States
transition_matrix <- matrix(c(0.1279, 0.8721,  
                              0.6631, 0.3369),  
                            nrow = 2, byrow = TRUE)

# Function to simulate a single Markov chain trial
simulate_markov_chain <- function(nSteps, transition_matrix, trial) {
  current_state <- "V"  
  sequence <- tibble(Step = 1, State = current_state, Trial = trial)
  
  for (i in 2:nSteps) {
    current_state <- if (current_state == "V") {
      sample(states, size = 1, prob = transition_matrix[1, ])
    } else {
      sample(states, size = 1, prob = transition_matrix[2, ])
    }
    sequence <- bind_rows(sequence, tibble(Step = i, State = current_state, Trial = trial))
  }
  
  sequence <- sequence %>% mutate(StateNum = ifelse(State == "V", 1, 2))
  return(sequence)
}

# Run multiple simulations
simulations <- bind_rows(lapply(1:nTrials, function(trial) {
  simulate_markov_chain(nSteps, transition_matrix, trial)
}))

# Plot the state transitions across trials as line graphs
ggplot(simulations, aes(x = Step, y = StateNum, group = Trial, color = State)) +
  geom_line(alpha = 0.5, size = 1) +  # Line graph with transparent lines
  scale_color_manual(values = c("V" = "blue", "C" = "red")) +
  scale_y_continuous(breaks = c(1, 2), labels = c("V", "C")) +
  theme_minimal() +
  labs(title = "Markov Chain State Transitions Across Multiple Trials",
       subtitle = "Each line represents a trial; Blue = 'V', Red = 'C'",
       x = "Time Step",
       y = "State") +
  theme(legend.position = "top")  # Move the legend to the top
```

QUESTION 4

The Markov model for the dolphi activity, long term distribution.
```{r}

# Define the coefficient matrix A
A <- matrix(c(
  0.84 - 1, 0.03, 0.01, 0.03, 0.03,
  0.11, 0.80 - 1, 0.15, 0.19, 0.09,
  0.01, 0.04, 0.70 - 1, 0.02, 0.05,
  0.04, 0.10, 0.07, 0.75 - 1, 0.00,
  1, 1, 1, 1, 1
), nrow = 5, byrow = TRUE)

# Define the right-hand side vector B
B <- c(0, 0, 0, 0, 1)

# Solve for pi using the solve function
pi_values <- solve(A, B)

# Print the solution
print(pi_values)
```

QUESTION 5

Moving from average to Rich in N steps.

N-STEP STOCHASTIC MATRICES


```{r}
# Define the transition matrix P
P <- matrix(c(0, 0.75, 0.2, 0.05,
              0.05, 0.2, 0.3, 0.45,
              0.1, 0.4, 0.3, 0.2,
              0, 0.15, 0.3, 0.55), 
            nrow = 4, byrow = TRUE)

# Function to calculate the n-step transition matrix
calculate_n_step_matrix <- function(P, n) {
  return(P %^% n)
}

# Calculate the 1-step, 2-step transition matrices
P_1_step <- calculate_n_step_matrix(P, 1)

# Calculate the 2-step transition matrices
P_2_step <- calculate_n_step_matrix(P, 2)

# Calculate the 3-step transition matrices
P_3_step <- calculate_n_step_matrix(P, 3)

# Print the results
cat("1-step transition matrix:\n")
print(P_1_step)

cat("\n2-step transition matrix:\n")
print(P_2_step)

cat("\n3-step transition matrix:\n")
print(P_3_step)

```

